Wildcard: Spreadsheet-Driven Customization of Web Applications

### 2020-01-14 paper.tex
submission ready

% Please always include the abstract.
% The abstract MUST be written according to the directives stated in 
% http://programming-journal.org/submission/
% Failure to adhere to the abstract directives may result in the paper
% being returned to the authors.
\begin{abstract}
Many Web applications do not meet the particular needs of their users.
Browser extensions offer a way to customize web applications, but most
people do not have the programming skills to implement their own
extensions.

We present the idea of \emph{spreadsheet-driven customization}: enabling
end users to customize existing applications using a live spreadsheet
view of the data inside the application. By manipulating the
spreadsheet, users can implement a variety of customizations without
doing any traditional programming.

We have implemented this technique in a prototype browser extension
called Wildcard. Through concrete examples, we demonstrate that Wildcard
can support useful customizations---ranging from sorting lists of search
results to showing related data from web APIs---on top of existing
websites. We also present the design principles underlying the
prototype.

Enabling easier customization can dramatically improve people's
experiences with software. We think that spreadsheet-driven
customization offers a promising new approach to the problem.

This paper is best viewed in the online version, which includes videos demonstrating the interface: \url{https://www.geoffreylitt.com/wildcard/salon2020}
\end{abstract}

@@

Web applications often don't match the particular needs of their users.
Sometimes there is a browser extension available to patch an issue, and
if the user is a programmer they might be able to fix it themselves. But
for most people, the only recourse is to complain to the developers, or
more likely, to simply give up. Back in 1977, in \emph{Personal Dynamic
Media} \autocite{kay1977}, Alan Kay envisioned personal computing as a
medium that let a user ``mold and channel its power to his own needs,''
but today, software behaves more like concrete than clay.

@@

We have implemented this technique in a prototype browser extension
called Wildcard. We've built demos of using Wildcard (shown in
Section~\ref{sec:demos}) which suggest that this paradigm can support
useful customizations, ranging from sorting lists of data to adding
whole new features to applications.

@@

Our approach requires extracting structured data from the user
interfaces of existing applications, but we hide the complexity of data
extraction from end users. Programmers write \emph{site adapters} which
use web scraping techniques to extract structured data from existing
applications and map them to the spreadsheet table. Our prototype
suggests that it is possible to implement site adapters for real
websites; Section~\ref{sec:implementation} describes some of the
techniques and challenges involved.

Spreadsheet-driven customization provides an easy entry point for end
users, since small tweaks like sorting data can be performed with a
single click. At the same time, it also supports a variety of richer
customizations, like adding private annotations to a webpage or joining
in related data from a web API. In Section~\ref{sec:design-principles},
we elaborate on this breadth of use cases, as well as other design
principles guiding our work.

Prior work \autocite{mccutchen2016,benson2014,chang2014} has enabled end
users to create ``spreadsheet-driven applications'' which use
spreadsheets as a backing data layer. Spreadsheet-driven
\emph{customization} applies this idea in a different context:
customizing existing software, rather than building new software from
scratch. Our technique does not require that the application actually be
backed by a spreadsheet; it merely uses the spreadsheet as an interface
for viewing and modifying the internal state of the application. In
Section~\ref{sec:related-work}, we describe how Wildcard relates to
existing work on spreadsheet-driven apps, as well as other areas like
web customization and web scraping.

The Wildcard extension is currently an early research prototype. We plan
to continue testing the system with our own use cases to explore how the
spreadsheet abstraction maps to real websites and customization needs.
Eventually we plan to release the tool publicly in order to study real
use cases, discover usability challenges, and to test the feasibility of
programmers building and maintaining site adapters.

@@

\hypertarget{customizing-search-results}{%
\subsection{Customizing search
results}\label{customizing-search-results}}

@@

Using Wildcard, the user can fix this omission, while leaving the page's
design and the rest of its functionality unchanged.{
Figure~\ref{fig:airbnb-demo} shows an overview of augmenting the Airbnb
site.} First, the user opens up the Wildcard panel, which shows a table
corresponding to the search results in the page. As they click around in
the table, the corresponding row in the page is highlighted so they can
see the connection between the views.

@@

After manipulating the data, the user closes the table view and
continues using the website with its original visual design. The table
view offers a way to change the data backing a page, but does not need
to replace the original interface entirely.

Most websites that show lists of data also offer actions on rows in the
table, like adding an item to a shopping cart. Wildcard has the ability
to make these ``row actions'' available in the data table through the
site adapter. In the Airbnb UI, saving multiple listings to a Favorites
list requires tediously clicking through them one by one. Using Wildcard
row actions, the user can select multiple rows and favorite all of them
with a single click{ (Figure~\ref{fig:airbnb-demo}, Note B)}. Similarly,
the user can open the detailed pages for many listings at once.

Next, the user wants to jot down some notes about each listing. To do
this, they type some notes into an additional column in each listing
row, and the notes appear inside the listings in the original UI{
(Figure~\ref{fig:airbnb-demo}, Note C)}. The annotations are saved in
the browser and associated with the ID of the listing, so they will
appear in future sessions that display the same listing.

Wildcard also includes a formula language that enables more
sophisticated customizations. When traveling without a car, it's useful
to evaluate potential places to stay based on how walkable the
surroundings are. Using a formula, the user can integrate Airbnb with
Walkscore, an API that rates the walkability of any location on a 1-100
scale. When the user uses the \texttt{walkscore} function with the
listing's latitude and longitude as inputs, it returns the walk score
for that location and shows it as the cell value. Because the cell's
contents are injected into the page, the score also appears in the UI{
(Figure~\ref{fig:airbnb-demo}, Note D)}.

@@

The user opens the table view, which shows the text and completed status
of each todo, and updates live in response to changes. They start the
customization by adding a new \texttt{snoozeDate} column to store the
snooze date for each todo{ (Figure~\ref{fig:todomvc-demo}, Note A)}.

The next step is to hide snoozed todos. The user creates a new
\texttt{snoozed?} column, which uses a formula to compute whether a todo
has a snooze date in the future{ (Figure~\ref{fig:todomvc-demo}, Note
B)}. Then, they simply filter the table to hide the snoozed todos{
(Figure~\ref{fig:todomvc-demo}, Note C)}.

As time passes, the built-in \texttt{NOW()} function will return the
current date, and snoozed todos will automatically appear at the right
time.

Because this implementation of snoozing was built on top of the
spreadsheet abstraction, it is completely decoupled from this particular
todo list app. We envision that users could package and share these
types of customizations as generic browser extensions, which could then
be instantly applied to any site supported by Wildcard.

\hypertarget{customizing-a-form-input}{%
\subsection{Customizing a form input}\label{customizing-a-form-input}}

@@

In this section we've presented just a few use cases for
spreadsheet-driven customization, to hint at the possibilities of the
paradigm. We think the spreadsheet model is flexible enough to support a
wide range of other useful modifications, while remaining familiar and
easy to use.

@@

\begin{figure}
\hypertarget{fig:architecture}{%
\centering
\includegraphics{media/architecture-clean.png}
\caption{The architecture of the Wildcard
prototype}\label{fig:architecture}
}
\end{figure}

@@

  \CommentTok{// Function to extract DOM element:}
\NormalTok{  el}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \VariableTok{row}\OperatorTok{.}\FunctionTok{querySelector}\NormalTok{(}\VerbatimStringTok{\textasciigrave{}.}\SpecialCharTok{$\{}\NormalTok{titleClass}\SpecialCharTok{\}}\VerbatimStringTok{\textasciigrave{}}\NormalTok{)}\OperatorTok{,}
\OperatorTok{\}}
\end{Highlighting}
\end{Shaded}

@@

First, the site adapter also needs to support sending updates from the
table to the original page. We have made this fairly straightforward,
because most DOM manipulation is not performed directly by the site
adapter---instead, Wildcard automatically mutates the DOM to reflect the
spreadsheet state, using the same declarative spec shown above. The only
exception to this is row actions (like favoriting an Airbnb listing),
which are simply implemented as Javascript functions with the freedom to
perform arbitrary behaviors like simulating clicks on buttons and
mutating the DOM.

Another challenge is triggering updates to the spreadsheet data in
response to UI changes that happen after initial page load. Site
adapters are responsible for implementing site-specific logic for
recognizing these changes only by observing the DOM. So far, we have
been able to use event listeners and the MutationObserver API to
successfully observe changes, but it may prove challenging to correctly
respond to updates on some sites only by observing the DOM.

@@

Computing does not need to be organized this way. For example, UNIX
offers a compelling alternative design: many small single-purpose
programs, all of which manipulate a universal format of text streams.
The universal format creates a high degree of leverage from tools: users
benefit from deeply mastering a text editor and other text manipulation
tools, since they can be applied to a huge variety of tasks. A user's
preferred text editor can even serve as an interactive input mechanism
in shell programs, e.g.~for editing git commit messages.

Spreadsheet-driven customization aims to bring some of the UNIX ethos to
the world of isolated Web applications, by creating a consistent data
structure to represent the data inside many applications. In UNIX, the
universal format is a text stream; in Wildcard, it is a relational
table. Because Wildcard maps the data from all applications to the table
format, users can invest in mastering the Wildcard table editor, the
formula language, and cell editor UIs, and reuse those same tools to
customize many different applications.

This idea relates to Beaudouin-Lafon and Mackay's notion of
\emph{polymorphic interface instruments} \autocite{beaudouin-lafon2000},
which are UI elements that can be used in different contexts (for
example, a color picker that can be used in many different drawing
applications). It also relates to ideas of literacy in a medium. DiSessa
notes that textual literacy rests on the fact that writing can be
adapted to many different genres and uses \autocite{disessa2000}; if
people needed to relearn reading and writing from scratch when switching
from essays to emails, the medium would lose most of its potency. Having
generic tools is especially important for software customization,
because the most common barrier to customizing software is not having
enough time \autocite{mackay1991}. It's more likely that people will
customize software regularly if they can reuse the same tools across
many applications.

This design principle leads to several challenges. First, any universal
abstraction has its constraints, and may not naturally express the data
in every application. We plan to explore the limits of the table
abstraction further, by trying to build adapters for more sites with
varied data formats. We expect that many types of data can fit easily
into tables: lists of search results, news articles, and messages can
all naturally be seen as relations. For sites that use document
structures (e.g.~Google Docs) or graph structures (e.g.~social
networks), it may prove more challenging to map internal data to the
table abstraction.

@@

One of the most interesting properties of spreadsheets is that users
familiar with only a tiny sliver of their functionality (e.g., storing
tables of numbers and computing simple sums) can still use them in
genuinely valuable ways. This supports the user's natural motivation to
continue using the tool, and to eventually learn its more powerful
features if needed \autocite{nardi1991}. In contrast, many traditional
programming systems require an enormous upfront time investment before
someone is able to write a program that helps them achieve a useful task
in their life.

As part of ensuring a low floor, we have focused on including features
that have immediate value for novices. For example, a user can sort a
table with a single click, or simply type in some annotations. We would
expect many Wildcard users to start by using these simpler features
before potentially moving on to more sophisticated features like
formulas.

Another aspect of providing a low floor is providing an ``in-place
toolchain'' \autocite{2019}: minimizing the effort of moving from using
to customizing, by making customization tools available in the same
environment where the user is already using the software. This quality
is distinct from the level of technical skill needed to use the tool.
For example, setting up a workflow trigger in an end user programming
system like \href{https://ifttt.com/}{IFTTT} does not require much
technical skill, but does require leaving the user's normal software and
entering a separate environment; conversely, running a Javascript
snippet in the browser console requires programming skills, but can be
done immediately and casually in the flow of using a website.

@@

Since we have still only built a few site adapters and demos, it is
still too early to tell exactly how high the ceiling is for the
customizations that can be achieved with Wildcard. But we think that
with enough built-in formulas (including standard spreadsheet formulas
and formulas for fetching data from other sources), the language could
support a wide variety of customizations. We plan to explore this aspect
further by trying to solve more real problems with the system and
observing where limitations emerge in practice.

@@

The first party developers of the original software can also play a role
in enabling customization. Spreadsheet-driven customization does not
depend on cooperation from first-party website developers, but if they
were to expose structured data in their web application clients, it
would eliminate the need for third party site adapters. We think there
are compelling reasons for first parties to consider doing this.
Providing Wildcard support would allow users to build extensions to
fulfill their own feature requests. It also would not necessarily
require much effort, since adding Wildcard support could be fairly
straightforward for a first-party that already has direct access to the
structured data in the page. There is also precedent for first parties
implementing an official client extension API: for several years, Google
maintained an official extension API in Gmail for Greasemonkey scripts
to use.

Google no longer maintains its own API, but third parties have continued
to maintain successful replacements---Gmail.js \autocite{talwar2019} is
an open source wrapper API with 77 contributors and hundreds of commits.
This shows another benefit of designing for an ecosystem: it's easier
for many programmers to collectively maintain a single adapter for a
site and keep it updated in response to changes, instead of creating
custom scraping logic for individual extensions.

@@

\begin{itemize}
\tightlist
\item
  Wildcard supports a broad set of customizations, but Sifter only
  supports sorting and filtering.
\item
  Wildcard shows the structured data table directly to the user. Sifter
  only shows sorting controls, and does not reveal the underlying data
  table.
\item
  Wildcard hides data extraction from end users. Sifter uses a
  combination of automated heuristics and interactive user feedback to
  involve end users in extracting data.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  \emph{Reliability}: It can be difficult to know whether a command will
  consistently work over time. In addition to changes to the website,
  changes to the heuristics can also cause problems.
\item
  \emph{Discoverability}: it can be difficult for users to discover the
  space of possible commands.
\end{itemize}

Wildcard offers a sharp contrast to sloppy programming, instead choosing
to expose a high degree of structure through the familiar spreadsheet
table. Wildcard offers more consistency; for example, clicking a sort
header will always work correctly as long as the site adapter is
maintained. Wildcard also offers clearer affordances for what types of
actions are possible, or, crucially, what actions are \emph{not}
possible, which is useful to know. On the other hand, Wildcard's
explicit site adapter approach means that fewer websites can be
customized, and also that users can't perform customizations if the
relevant data is not in the spreadsheet.

@@

In the future, we might try to integrate existing web scraping tools, to
help create more reliable site adapters for Wildcard with less work, and
to open up adapter creation to end users.

\hypertarget{future-work}{%
\section{Future Work}\label{future-work}}

There are still many open questions about spreadsheet-driven
customization which we hope to answer through targeted development and
usage of the Wildcard prototype.

@@

We suspect that two areas of the current model may prove limiting.
First, Wildcard's data model currently shows a single table at a time,
without any notion of relationships between tables. A richer data model
with foreign keys might help support certain use cases. For designing a
spreadsheet interface on top of a richer relational model, we could
learn from other systems with this design
\autocite{mccutchen2016,bakke2016}.

A second limitation is that there is currently no way for end users to
express imperative workflows with sequences of actions. Related work has
shown these types of workflows to be useful
\autocite{bolin2005,leshed2008} in practice. It's not clear how such
workflows could fit into Wildcard, since spreadsheets have a
fundamentally different computation model, and site adapters created by
programmers cannot easily account for every possible action a user would
want to take in a page.

@@

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

We plan to continue developing the Wildcard prototype and to eventually
deploy it as an open-source tool. To receive future updates on Wildcard
and notifications about a public release,
\href{https://geoffreylitt.substack.com/welcome}{sign up for the email
newsletter}.

We are also looking for private beta testers. If you have an idea for
how you might want to use Wildcard, please
\href{mailto:glitt@mit.edu}{get in touch}. We would love to hear about
your needs and help find ways to use Wildcard to solve them.

>>>>

% Please always include the abstract.
% The abstract MUST be written according to the directives stated in 
% http://programming-journal.org/submission/
% Failure to adhere to the abstract directives may result in the paper
% being returned to the authors.
\begin{abstract}
Many Web applications do not meet the precise needs of their users.
Browser extensions offer a way to customize web applications, but most
people do not have the programming skills to implement their own
extensions.

We have implemented this technique in a prototype browser extension
called Wildcard. Through concrete examples, we demonstrate that Wildcard
can support useful customizations---ranging from sorting lists of search
results to showing related data from web APIs---on top of existing
websites. We also present the design principles underlying our
prototype.

Customization can lead to dramatically better experiences with software.
We think that spreadsheet-driven customization offers a promising new
approach to unlocking this benefit for all users, not just programmers.

This paper is best viewed in the online version (\url{https://www.geoffreylitt.com/wildcard/salon2020}), which includes videos demonstrating the interface.
\end{abstract}

@@

Web applications often don't meet the precise needs of their users.
Sometimes there is a browser extension available to patch an issue, and
if the user is a programmer they might be able to fix it themselves. But
for most people, the only recourse is to complain to the developers, or
more likely, to simply give up. Back in 1977, in \emph{Personal Dynamic
Media} \autocite{kay1977}, Alan Kay envisioned personal computing as a
medium that let a user ``mold and channel its power to his own needs,''
but today, software behaves more like concrete than clay.

@@

We have implemented this technique in a prototype browser extension
called Wildcard, and used it to build demos which suggest that this
paradigm can support useful customizations, ranging from sorting lists
of data to adding whole new features to applications (shown in
Section~\ref{sec:demos}).

@@

Our approach requires extracting structured data from the user
interfaces of existing applications, but we hide the complexity of data
extraction from end users. Programmers write \emph{site adapters} which
use web scraping techniques to extract structured data from existing
applications and map it to the spreadsheet table. Our prototype suggests
that it is possible to implement site adapters for real websites;
Section~\ref{sec:implementation} describes some of the techniques and
challenges involved.

Spreadsheet-driven customization is based on three design principles,
described in Section~\ref{sec:design-principles}:

\begin{itemize}
\tightlist
\item
  \textbf{Expose a universal structure}: By mapping the data in all
  applications to the same universal table abstraction, Wildcard enables
  users to learn a generic customization toolkit that they can apply to
  any site.
\item
  \textbf{Low floor, high ceiling}: Wildcard provides an easy entry
  point for end users, since small tweaks like sorting data can be
  performed with a single click. At the same time, it also supports a
  variety of richer customizations, like adding private annotations to a
  webpage or joining in related data from a web API.
\item
  \textbf{Design for an ecosystem of users}: Spreadsheet-driven
  customization combines the efforts of programmers and end users,
  rather than putting the full burden on end users. We also envision a
  role for first party developers to help make their applications more
  customizable.
\end{itemize}

Prior work \autocite{mccutchen2016,benson2014,chang2014} has enabled end
users to create ``spreadsheet-driven applications'' which use
spreadsheets as a backing data layer. Spreadsheet-driven
\emph{customization} applies this idea in a different context:
customizing existing software, rather than building new software from
scratch. Our technique does not require that the application actually be
backed by a spreadsheet; it merely uses the spreadsheet as an interface
for viewing and modifying the internal state of the application. In
Section~\ref{sec:related-work}, we describe further how Wildcard relates
to existing work on spreadsheet-driven apps, as well as work in other
areas like web customization and web scraping.

The Wildcard extension is currently an early research prototype. We plan
to continue testing the system with our own use cases to explore how the
spreadsheet abstraction maps to real websites and customization needs.
Eventually we plan to release the tool publicly in order to learn from
real use cases, discover usability challenges, and to test the
feasibility of building and maintaining site adapters.
Section~\ref{sec:conclusion} describes open questions we plan to
continue exploring, as well as our future plans.

@@

\hypertarget{augmenting-search-results}{%
\subsection{Augmenting search results}\label{augmenting-search-results}}

@@

Using Wildcard, the user can fix this omission, while leaving the page's
design and the rest of its functionality unchanged.{
Figure~\ref{fig:airbnb-demo} shows an overview of augmenting the Airbnb
site.} First, the user opens up the Wildcard panel, which shows a table
corresponding to the search results in the page. As they click around in
the table, the corresponding row in the page is highlighted to indicate
the mapping between the views.

@@

After manipulating the data, the user closes the table view and
continues using the website. Because the application's UI usually has a
nicer visual design than a spreadsheet, Wildcard does not aim to replace
it---at any time, the user can use either the UI, the spreadsheet, or
both together.

Many websites that show lists of data also offer actions on rows in the
table, like adding an item to a shopping cart. Wildcard has the ability
to make these ``row actions'' available in the data table through the
site adapter. In the Airbnb UI, saving multiple listings to a Favorites
list requires tediously clicking through them one by one. Using Wildcard
row actions, the user can select multiple rows and favorite all of them
with a single click{ (Figure~\ref{fig:airbnb-demo}, Note B)}. Similarly,
the user can open the detailed pages for many listings at once.

Next, the user wants to jot down some notes about each listing. To do
this, they type some notes into an additional column in each row, and
the notes appear inside the listings in the original UI{
(Figure~\ref{fig:airbnb-demo}, Note C)}. The annotations are saved in
the browser and associated with the backend ID of the listing, so they
will appear in future browser sessions that display the same listing.

Wildcard also includes a formula language that enables more
sophisticated customizations. When traveling without a car, it's useful
to evaluate potential places to stay based on how walkable the
surroundings are. Using a formula, the user can integrate Airbnb with
Walkscore, an API that rates the walkability of any location on a 1-100
scale. When the user calls the \texttt{walkscore} formula with the
listing's latitude and longitude as arguments, it returns the walk score
for that location and shows it as the cell value. Because the cell's
contents are injected into the page, the score also appears in the UI{
(Figure~\ref{fig:airbnb-demo}, Note D)}.

@@

The user opens the table view, which shows the text and completed status
of each todo. They start the customization by adding a new column to
store the snooze date for each todo{ (Figure~\ref{fig:todomvc-demo},
Note A)}.

The next step is to hide snoozed todos. The user creates a
\texttt{snoozed?} column, which uses a formula to compute whether a todo
is snoozed---i.e., whether it has a snooze date in the future{
(Figure~\ref{fig:todomvc-demo}, Note B)}. Then, they simply filter the
table to hide the snoozed todos{ (Figure~\ref{fig:todomvc-demo}, Note
C)}.

Because the built-in \texttt{NOW()} function always returns the current
datetime, snoozed todos will automatically appear once their snooze date
arrives.

Because this implementation of snoozing was built on the spreadsheet
abstraction, it is completely decoupled from this particular todo list
app. We envision that users could share these types of customizations as
generic browser extensions, which could be applied to any site supported
by Wildcard with no additional effort.

\hypertarget{adding-a-custom-datepicker}{%
\subsection{Adding a custom
datepicker}\label{adding-a-custom-datepicker}}

@@

In this section we've presented just a few use cases for
spreadsheet-driven customization, to suggest some possibilities of the
paradigm. We think the spreadsheet model is flexible enough to support a
wide range of other useful modifications, while remaining familiar and
easy to use.

@@

\begin{figure}
\hypertarget{fig:architecture}{%
\centering
\includegraphics{media/architecture-300dpi.png}
\caption{The architecture of the Wildcard
prototype}\label{fig:architecture}
}
\end{figure}

@@

  \CommentTok{// Function to extract DOM element:}
\NormalTok{  el}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \VariableTok{row}\OperatorTok{.}\FunctionTok{querySelector}\NormalTok{(}\VerbatimStringTok{\textasciigrave{}.\_1ebt2xej\textasciigrave{}}\NormalTok{)}\OperatorTok{,}
\OperatorTok{\}}
\end{Highlighting}
\end{Shaded}

@@

First, the site adapter needs to support sending updates from the table
to the original page. We have made this fairly straightforward, because
most DOM manipulation is not performed directly by the site
adapter---instead, Wildcard automatically mutates the DOM to reflect the
spreadsheet state, using the same declarative spec shown above. The only
exception to this is row actions (like favoriting an Airbnb listing),
which are implemented as Javascript functions that can perform arbitrary
behaviors like simulating clicks on buttons or mutating the DOM.

Another challenge is triggering updates to the spreadsheet data in
response to UI changes that happen after initial page load. Site
adapters are responsible for implementing site-specific logic for
recognizing these changes by observing the DOM. So far, we have been
able to use event listeners and the MutationObserver API to successfully
observe changes, but it may prove challenging to correctly respond to
updates on some sites purely by observing the DOM.

@@

Computing does not need to be organized this way. For example, UNIX
offers a compelling alternative design: many small single-purpose
programs, all of which manipulate a universal format of text streams.
Users get high leverage from mastering a text editor and other text
manipulation tools, since they can be applied to a huge variety of
tasks. A user's preferred text editor can even serve as an interactive
input mechanism in shell programs, e.g.~for editing git commit messages.

Spreadsheet-driven customization aims to bring some of the UNIX ethos to
the world of isolated Web applications, by creating a consistent data
structure to represent the data inside many applications. In UNIX, the
universal format is a text stream; in Wildcard, it is a relational
table. Because Wildcard maps the data from all applications to the table
format, users can master the tools provided by Wildcard and apply them
to customize many different applications.

This idea relates to Beaudouin-Lafon and Mackay's notion of
\emph{polymorphic interface instruments} \autocite{beaudouin-lafon2000},
which are UI elements that can be used in different contexts (for
example, a color picker that can be used in many different drawing
applications). It also relates to ideas of literacy in a medium. DiSessa
notes that textual literacy depends on the fact that writing can be
adapted to many different genres and uses \autocite{disessa2000}; if
people needed to relearn reading and writing from scratch when switching
from essays to emails, the medium would lose its potency. Having generic
tools is especially important for software customization, because a
common barrier to customizing software is not having enough time
\autocite{mackay1991}. It's more likely that people will customize
software regularly if they can reuse the same tools across many
applications.

This design principle leads to several challenges. First, any universal
abstraction has its constraints, and may not naturally express the data
in every application. We plan to explore the limits of the table
abstraction further, by trying to build adapters for more sites with
varied data formats. We expect that many types of data will fit easily
into tables: lists of search results, news articles, and messages can
all naturally be seen as relations. Documents (e.g.~Google Docs) or
graphs (e.g.~social network friend graphs) may prove more challenging to
map to the table abstraction.

@@

One of the most interesting properties of spreadsheets is that users
familiar with only a tiny sliver of their functionality (e.g., storing
tables of numbers and computing simple sums) can still use them in
genuinely valuable ways. This supports the user's natural motivation to
continue using the tool, and to eventually learn its more powerful
features if needed \autocite{nardi1991}. In contrast, many traditional
programming systems require an enormous upfront time investment before
someone is able to write a program that helps them achieve a useful
task.

As part of ensuring a low floor, we have focused on including features
that have immediate value for novices. For example, a user can sort a
table with a single click, or simply type in some annotations. We would
expect many Wildcard users to start by using these simpler features
before moving on to more sophisticated features like formulas.

Another aspect of providing a low floor is providing an ``in-place
toolchain'' \autocite{2019}: minimizing the effort of moving from using
to customizing, by making customization tools available in the same
environment where the user is already using the software. This quality
is distinct from the level of technical skill needed to use the tool.
For example, setting up a workflow in an end user programming system
like \href{https://ifttt.com/}{IFTTT} does not require much programming
skill, but it does require leaving the user's normal software and
entering a separate environment; conversely, running a Javascript
snippet in the browser console requires programming skills, but can be
done immediately in the flow of using a website.

@@

Since we have still only built a few site adapters and demos, it is
still too early to tell how high the ceiling is for the customizations
that can be achieved with Wildcard. But we think that with enough
built-in formulas (both standard spreadsheet formulas for computation,
and formulas for fetching data from other sources), the language could
support a wide variety of customizations. We plan to explore this aspect
further by trying to solve real problems and observing where limitations
emerge in practice. Section~\ref{sec:demos} also discusses some
limitations we expect to face.

@@

The first party developers of the original software can also play a role
in enabling customization. Spreadsheet-driven customization does not
depend on cooperation from first-party website developers, but if they
were to expose structured data in their web application clients, it
would eliminate the need for third party site adapters. We think there
are compelling reasons for first parties to consider doing this.
Providing Wildcard support would allow users to build extensions to
fulfill their own feature requests. It also would not necessarily
require much effort, since adding Wildcard support could be fairly
straightforward for a first-party that already has direct access to the
structured data in the page. (It would certainly be far simpler than
implementing a third-party site adapter.) There is also precedent for
first parties implementing an official client extension API: for several
years, Google maintained an official extension API in Gmail for
Greasemonkey scripts to use.

Google no longer maintains its own API, but third parties have continued
to maintain successful replacements---Gmail.js \autocite{talwar2019} is
an open source wrapper API with over 70 contributors and hundreds of
commits. This library demonstrates another benefit of designing for an
ecosystem: it's easier for many programmers to collectively maintain a
single adapter for a site and keep it updated in response to changes to
the site, rather than each programmer creating custom scraping logic for
their own individual extensions.

@@

\begin{itemize}
\tightlist
\item
  Wildcard hides data extraction from end users by having programmers
  develop site adapters, whereas Sifter uses a combination of automated
  heuristics and interactive user feedback to involve end users in
  extracting data. User studies of Sifter indicated that users were
  surprised by the need to participate in the data augmentation process,
  which influenced our decision to hide this process from the user. We
  suspect Wildcard's workflow results in a simpler end user experience,
  but at the cost of only working with a smaller subset of supported
  websites.
\item
  Wildcard supports a broad set of customizations. Sifter only supports
  sorting and filtering.
\item
  Wildcard shows the structured data table directly to the user. Sifter
  only shows sorting controls, and does not reveal the underlying data
  table.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  \emph{Reliability}: It can be difficult to know whether a command will
  consistently work over time. In addition to changes to the website,
  changes to the heuristics can also cause problems.
\item
  \emph{Discoverability}: it can be hard for users to find discover
  possible commands.
\end{itemize}

Wildcard offers a sharp contrast to sloppy programming, instead choosing
to expose a high degree of structure through the spreadsheet table.
Wildcard offers more consistency; for example, clicking a sort header
will always work correctly as long as the site adapter is maintained.
Wildcard also offers clearer affordances for what types of actions are
possible. On the other hand, Wildcard's explicit site adapter approach
means that fewer websites can be customized, and also that users can't
perform customizations if the relevant data is not exposed in the
spreadsheet.

Another difference between Wildcard and these tools is that there is no
way for end users to express imperative workflows with sequences of
actions in Wildcard. Chickenfoot and Coscripter have shown these types
of workflows to be useful in practice, but it's not clear how such
workflows could fit into Wildcard. Spreadsheets have a fundamentally
different computation model, and site adapters created by programmers
cannot easily account for every possible action a user might want to
perform in a page.

@@

In the future, we might try to integrate existing web scraping tools, to
help create more reliable site adapters for Wildcard with less work, and
to open up adapter creation to end users. Although Wildcard has
additional requirements beyond traditional web scraping as discussed in
Section~\ref{sec:implementation}, there is still enough overlap that
existing tools might prove helpful.

\hypertarget{sec:conclusion}{%
\section{Conclusion and Future Work}\label{sec:conclusion}}

There are still many open questions about spreadsheet-driven
customization which we hope to answer through targeted development and
usage of Wildcard.{\footnote{Up to date information about Wildcard is
  available on the project website
  (\url{https://geoffreylitt.com/wildcard}), which includes a signup for
  receiving updates and notifications about a public release. We also
  welcome hearing from readers who might be interested in being private
  beta testers or who have feedback on the project.}}

@@

We suspect that some areas of the current model may prove overly
simplistic. For example, Wildcard's data model currently shows a single
table at a time, without any notion of relationships between tables. A
richer data model with foreign keys might help support certain use
cases. For designing a spreadsheet interface on top of a richer
relational model, we could learn from other systems with this design
\autocite{mccutchen2016,bakke2016}.

@@
